---
title: "TINT - gene ratios"
author: "Eva Freyhult"
date:  "today"
date-format: long
embed-resources: true
format: 
  html:
    page-layout: full
    number-sections: true
    toc: true
  docx:
    number-sections: true
    toc: true
execute: 
  echo: false
  message: false
  warning: false
  cache: true
knitr:
  opts_chunk:
    fig.path: "figure/" 
    dev:
     - png
     - pdf
params:
  esetdir: '~/Support/4709/TINT/data/esets'
---

```{r setupratios}
library(tidyverse)
library(gt)
library(ggplot2)

library(oligo)

library(limma)
library(lmerTest)
library(multcomp)
library(broom)
```


```{r loaddata}
eset <- readRDS(file.path(params$esetdir, "esetTINT.Rds"))
pheno <- pData(eset)

##Storage
##All samples analyzed in 2021, patientid gives sample year
pheno <- pheno |> mutate(storagetime=21-as.numeric(sub(".*-", "", pheno$patientid)))

##Variance
fData(eset)$var <-  apply(exprs(eset), 1, var)

## Remove control probesets
eset <- eset[!grepl("^AFFX|^ERCC|pos|neg", rownames(eset)),]
## Keep only protein coding genes
fData(eset) <- fData(eset) %>% mutate(protein=grepl("^protein-coding$", GENETYPE))
esetp <- eset[fData(eset)$protein,]
```

## Description

Investigate expression of activitry markers for CD3, CD8 and CD68 positive cells between high and low grade tumors.

Compute gene expression ratios and compare these ratios (or rather the log2 ratios) between ISUP 3-5 and 1-2 in T, TINT-N and TINT-D. The ratios to investigate are:

* PRF1, GZMB, GZMA, GNLY, CTLA-4, PD1, FAS, IFNG vs. CD3
* PRF1, GZMB vs. CD8
* CTLA-4 vs. CD4
* CD163, IDO1, MRC1 (CD206), SPP1, PDL1 (CD274),  CXCL9, CXCL10, S100A9 vs. CD68


## Methods

For CD3 compute the mean value over CD3D and CD3E.

For CD8 compute the mean value over CD8A and CD8B.

Compute log2 ratios based on the RMA normalized data as the differences between log2 intensities.

Associations between ISUP groups (3-5 vs 1-2) and log2-ratios are computed using linear mixed effects models with random intercept for patient. Two versions of the models are computed; one with only ISUP group, sample type (T, T1, T2) and the interaction ISUPgroup:sampletype and one that in addition include age and run.

```{r ratios}
fd <- fData(esetp) |> filter(SYMBOL %in% c("PRF1", "GZMB", "GZMA", "GNLY", "CTLA4", "PDCD1", "FAS", "IFNG", "CD3E", "CD3D", "CD8A", "CD8B", "CD4", "CD163", "IDO1", "MRC1", "SPP1", "CD274", "CXCL9", "CXCL10", "S100A9", "CD68", "CD3G"))

X0 <- exprs(esetp[fd$PROBEID,])
##Compute mean per SYMBOL
spl <- split(fd$PROBEID, fd$SYMBOL)
X0 <- sapply(spl, function(i) colMeans(X0[i, , drop=FALSE])) |> t()
rownames(X0) <- names(spl)

## Exclude CD3G AND compute mean per CD3 and CD8
spl$CD3G <- NULL
nms <- sub("CD8.*", "CD8", sub("CD3.*", "CD3", names(spl)))
spl <- split(names(spl), nms)
X <- sapply(spl, function(i) colMeans(X0[i, , drop=FALSE])) |> t()
rownames(X) <- names(spl)

pairs <- list(
  PRF1_CD3   = c("PRF1",  "CD3"),
  GZMB_CD3   = c("GZMB",  "CD3"),
  GZMA_CD3   = c("GZMA",  "CD3"),
  GNLY_CD3   = c("GNLY",  "CD3"),
  CTLA4_CD3  = c("CTLA4", "CD3"),
  PDCD1_CD3  = c("PDCD1", "CD3"),
  FAS_CD3    = c("FAS",   "CD3"),
  IFNG_CD3   = c("IFNG",  "CD3"),

  PRF1_CD8   = c("PRF1",  "CD8"),
  GZMB_CD8   = c("GZMB",  "CD8"),

  CTLA4_CD4  = c("CTLA4", "CD4"),

  CD163_CD68 = c("CD163", "CD68"),
  IDO1_CD68  = c("IDO1",  "CD68"),
  MRC1_CD68  = c("MRC1",  "CD68"),
  SPP1_CD68  = c("SPP1",  "CD68"),
  CD274_CD68 = c("CD274", "CD68"),
  CXCL9_CD68 = c("CXCL9", "CD68"),
  CXCL10_CD68= c("CXCL10","CD68"),
  S100A9_CD68= c("S100A9","CD68")
)

ratios <- sapply(pairs, function(g) {
  X[g[1], ] - X[g[2],]
})
```

## Results

### CD3 and CD8

CD3 is represented on the array by CD3D, CD3E and CD3G. CD8 is represented by CD8A and CD8B.

In the following analyses mean values will be used, for CD3 over CD3D and CD3E and for CD8 over CD8A and CD8B.

An overview of how similar (or not similar) the values are is shown in @fig-CDpairs.

```{r fig-CDpairs}
#| fig-cap: "Scatter plots comparing the CD3 and CD8 genes. Pearson correlation coefficient is printed."
library(GGally)
ggpairs(as.data.frame(t(X0)) |> dplyr::select(starts_with(c("CD3", "CD8"))) |> rownames_to_column("sampleid") |> left_join(pheno, by="sampleid"), columns = 2:6, mapping=aes(color=sample.type, alpha=0.5))
```

### Ratio associations

```{r lm}
pheno <- pheno |> mutate(ISUPgr = factor(ISUP, levels=1:5, labels=rep(c("1-2", "3-5"), c(2,3))))

ratiosl <- as.data.frame(ratios) |> rownames_to_column("sampleid") |> pivot_longer(-sampleid, names_to = "ratio") |> left_join(pheno, by="sampleid")

forms <- c(ISUPnocovars=~ISUPgr*sample.type,
           ISUPagerun=~ISUPgr*sample.type+age+run)      
designs <- lapply(forms, function(f) {
  des <- model.matrix(f, data=pheno)
  colnames(des) <- make.names(colnames(des))
  des
})

#dupcor <- duplicateCorrelation(X, design=designs$ISUPagerun, block=pheno$patientid)

contrastISUP <- makeContrasts(
  ISUP35vs12_T1 = ISUPgr3.5 + ISUPgr3.5.sample.typeT1,
  ISUP35vs12_T2 = ISUPgr3.5 + ISUPgr3.5.sample.typeT2,
  ISUP35vs12_T = ISUPgr3.5,
  levels=designs$ISUPagerun)

lmcontrast <- function(form, contrast, df) {
  f <- lmer(form, data=df)
  nm <- make.names(names(fixef(f)))
  tidy(glht(f, linfct=t(contrast[nm, ])), test=univariate(), conf.int=TRUE)
}

res <- lapply(forms, function(form) ratiosl |> group_by(ratio) |> reframe(lmcontrast(form=update(form, value ~ . + (1|patientid)), contrast=contrastISUP, pick(everything()))))
```

```{r tabs}
tabs <- lapply(res, function(r) r |> mutate(log2FC=sprintf("%.2f (%.2f, %.2f)", estimate, conf.low, conf.high), p=adj.p.value) |> pivot_wider(id_cols = c(ratio), names_from=contrast, values_from = c(log2FC, p), names_vary = "slowest") |> relocate(ends_with("T"), .after=ratio) |> gt() |> tab_spanner(label="3-5 vs 1-2, T", columns=ends_with("ISUP35vs12_T")) |> tab_spanner(label="3-5 vs 1-2, TINT1", columns=ends_with("ISUP35vs12_T1")) |> tab_spanner(label="3-5 vs 1-2, TINT2", columns=ends_with("ISUP35vs12_T2")) |> cols_label_with(fn=~sub("log2FC", "log2FC (95% CI)", gsub("(.*)_.*_.*", "\\1", .))) |> fmt_number(decimal=3) |> 
  tab_style(style = cell_fill(color = "#f4cccc"), locations = cells_body(columns = c(log2FC_ISUP35vs12_T1, p_ISUP35vs12_T1), rows = p_ISUP35vs12_T1 <= 0.05&log2FC_ISUP35vs12_T1>0)) |>
  tab_style(style = cell_fill(color = "#cfe2f3"), locations = cells_body(columns = c(log2FC_ISUP35vs12_T1, p_ISUP35vs12_T1), rows = p_ISUP35vs12_T1 <= 0.05&log2FC_ISUP35vs12_T1<0)) |> 
  tab_style(style = cell_fill(color = "#f4cccc"), locations = cells_body(columns = c(log2FC_ISUP35vs12_T2, p_ISUP35vs12_T2), rows = p_ISUP35vs12_T2 <= 0.05&log2FC_ISUP35vs12_T2>0)) |>
  tab_style(style = cell_fill(color = "#cfe2f3"), locations = cells_body(columns = c(log2FC_ISUP35vs12_T2, p_ISUP35vs12_T2), rows = p_ISUP35vs12_T2 <= 0.05&log2FC_ISUP35vs12_T2<0)) |> 
  tab_style(style = cell_fill(color = "#f4cccc"),locations = cells_body(columns = c(log2FC_ISUP35vs12_T, p_ISUP35vs12_T), rows = p_ISUP35vs12_T <= 0.05&log2FC_ISUP35vs12_T>0)) |>
  tab_style(style = cell_fill(color = "#cfe2f3"), locations = cells_body(columns = c(log2FC_ISUP35vs12_T, p_ISUP35vs12_T), rows = p_ISUP35vs12_T <= 0.05&log2FC_ISUP35vs12_T<0)) )
```

```{r tbl-rationocovars}
#| tbl-cap: "Association between ratios and ISUP 3-5 and 1-2. log2(fold change) with 95% confidence interval and p-value. No covariates in the model."
tabs$ISUPnocovars
```

```{r tbl-ratioagerun}
#| tbl-cap: "Association between ratios and ISUP 3-5 and 1-2. log2(fold change) with 95% confidence interval and p-value. Age and run are included as covariates in the model."
  tabs$ISUPagerun
```

